AI Safety for High Energy Physics

### 2019-10-04 documentation/manuscript.tex
worked through IV

After clustering, a jet $J_i$ is represented as a truncated list of $N_i\leq\nconst$ constituent 4-momenta $J_i = \{(\pt^k, \eta^k,\phi^k) \colon k = 1..N^i\}$\footnote{Using collider coordinates, $p_T$ is the particle momentum transverse to the collision axis, $\phi$ is the azimuthal angle and the pseudo-rapidity $\eta$ is $-\ln(\tan(\theta/2))$, where $\theta$ is the polar angle.}., corresponding to the calorimeter hits included in the jet, resulting in a data dimensionality of approximately 200.
The $\pt$ of each constituent is expressed in units of TeV, resulting in a dynamic range that is of order unity.
These jets are then uses as inputs to two different network architectures which are trained to discriminate signal jets from background.

@@

The first architecture, referred to as the \textit{High-level} (HL) model, is a simple fully-connected neural network, except that the first layer has no learnable parameters and instead explicitly calculates four jet-level observables from the input constituents: $\pt$, $\eta$, invariant mass, and $D_2$.
The first three features simply represent the 4-vector associated with the jet cluster; the quantity $D_2$~\cite{Larkoski:2014gra} is motivated by QCD theory and designed to discriminate jets with radiation patterns consistent with multiple sub-jet axes, characteristic of boosted boson decays.
The remainder of the network is comprised of $\HLLayers$ dense layers, each with $\HLUnits$ units and ReLU activation, followed by a single output neuron with sigmoid activation.

The second architecture, referred to as the \textit{Low-level} (LL) model, is a Particle Flow Network~\cite{Komiske:2018cqr}, which operates directly on the full list of input constituents.
Although we also observed very similar results using a simple fully-connected network, the PFN architecture has the property of invariance with respect to permutations of the input constituents, which simplifies the interpretation later on.
The benchmark LL model has $\PfnPhiLayers$ layers in the $\Phi$ subnetwork and $\PfnFLayers$ layers in the $F$ subnetwork, all with $\PfnUnits$ units and ReLU activations.
The final layer is again a single neuron with sigmoid activation.

@@

\subsection{Adversarial Network Method}
%%%% TODO: I have not looked at this at all since the first pass a few weeks ago, needs update -C
In this approach, we train a second neural network $g$ for which both input and output are $\nconst\times 3$ tensors.
The goal is to train $g$ to map $J\mapsto\delta J$, for an arbitrary jet such that the classifier network $f$ will perceive the input $J+\delta J$ as a background jet.
The network $g$ is trained by minimizing separate loss functions for signal and background defined by:
\begin{align}
\mathcal{L}_\text{sig}&=\log(1-f(g(J))),\\
\mathcal{L}_\text{bg}&=\lambda_\mathrm{cls} (f(J)-f(g(J)))^2+ \lambda^{(i)}_\mathrm{obs} (\mathcal{O}^{(i)}(J)-\mathcal{O}^{(i)}(g(J))^2,
\end{align}
where the functions $\mathcal{O}^{(i)}(J)$ are functions $\mathbb{R}^{3N}\rightarrow \mathbb{R}$, representing the one-dimensional observable features to be preserved for validation.
The tunable hyperparameters $\lambda_\mathrm{cls}, \lambda^{(i)}_\mathrm{obs} \geq 0$ encode the adversary's preference to preserve the target network response and observable features, respectively. for background events.

In our experiments, $g$ is a fully-connected network with \nlayerADV\ hidden layers, each with \nunitsADV\ units and ReLU activation.
The final output layer has $\nconst\times 3$ units, with $\tanh$ activations.
Analogously to the bounding parameters $\epsilon$ in \ref{subsec:fgsm}, the outputs of the final layer are bounded by applying a $\tanh$ activation functions and the axes corresponding to $\pt$, $\eta$, and $\phi$ are scaled by parameters $\rho_{\pt}$, $\rho_\eta$, and $\rho_\phi$ respectively.

@@

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.45\textwidth]{figures/FGSM_Significance.pdf}
\caption{Top: The relative discovery significance (true positive rate / square root of false positive rate) as a function of the classifier threshold for HL observables and LL observables before and after the FGSM attack described in Sec.~\ref{subsec:fgsm}.  Bottom: the $D_2$ observable before and after the FGSM attack for signal and background. }
\label{fig:FGSM2}
\end{figure}

>>>>

After clustering, a jet $J_i$ is represented as a truncated list of $N_i\leq\nconst$ constituent 4-momenta $J_i = \{(\pt^k, \eta^k,\phi^k) \colon k = 1,...,N_i\}$\footnote{Using collider coordinates, $\pt$ is the particle momentum transverse to the collision axis, $\phi$ is the azimuthal angle and the pseudo-rapidity $\eta$ is $-\ln(\tan(\theta/2))$, where $\theta$ is the polar angle.}.  The resulting dimensionality of the features used for learning is about 200.
The $\pt$ of each constituent is expressed in units of TeV, resulting in a dynamic range that is of order unity.
These jets are then uses as inputs to two different network architectures which are trained to discriminate signal jets from background jets.

@@

The first architecture, referred to as the \textit{High-level} (HL) model, is a simple fully-connected neural network, except that the first layer has no learnable parameters and instead explicitly calculates four jet-level observables from the input constituents: $\pt$, $\eta$, invariant mass, and $D_2$.
The first three features simply represent the 4-vector associated with the entire jet; the quantity $D_2$~\cite{Larkoski:2014gra} is theoretically motivated from the strong force and designed to identify jets with radiation patterns consistent with two sub-jet axes, characteristic of Lorentz boosted boson decays.
The remainder of the network is comprised of $\HLLayers$ dense layers, each with $\HLUnits$ units and ReLU activation, followed by a single output neuron with sigmoid activation.

The second architecture, referred to as the \textit{Low-level} (LL) model, is a Particle Flow Network~\cite{Komiske:2018cqr}, which operates directly on the full list of input constituents.
Although we also observed very similar results using a simple fully-connected network, the PFN architecture has the property of invariance with respect to permutations of the input constituents, which simplifies the interpretation as order perturbations have no impact on the result.
The benchmark LL model has $\PfnPhiLayers$ layers in the $\Phi$ subnetwork and $\PfnFLayers$ layers in the $F$ subnetwork, all with $\PfnUnits$ units and ReLU activations.
The final layer is again a single neuron with sigmoid activation.

@@

\subsection{Adversarial Network Method}
%%%% TODO: I have not looked at this at all since the first pass a few weeks ago, needs update -C
In this approach, we train a second neural network $g$ for which both input and output are $\nconst\times 3$ tensors.
The goal is to train $g$ to map $J\mapsto\delta J$, for an arbitrary jet such that the classifier network $f$ will perceive the input $J+\delta J$ as a background jet.
The network $g$ is trained by minimizing separate loss functions for signal and background defined by:
\begin{align}
\mathcal{L}_\text{sig}&=\log(1-f(g(J))),\\\nonumber
\mathcal{L}_\text{bg}&=\lambda_\mathrm{cls} (f(J)-f(g(J)))^2\\
&\hspace{3mm}+\sum_i \lambda^{(i)}_\mathrm{obs} (\mathcal{O}^{(i)}(J)-\mathcal{O}^{(i)}(g(J))^2,
\end{align}
where the functions $\mathcal{O}^{(i)}(J)$ are functions $\mathbb{R}^{3N}\rightarrow \mathbb{R}$, representing the one-dimensional observable features to be preserved for validation.
The tunable hyperparameters $\lambda_\mathrm{cls}, \lambda^{(i)}_\mathrm{obs} \geq 0$ encode the adversary's preference to preserve the target network response and observable features, respectively, for background events.

In our experiments, $g$ is a fully-connected network with \nlayerADV\ hidden layers, each with \nunitsADV\ units and ReLU activation.
The final output layer has $\nconst\times 3$ units, with $\tanh$ activations.
Analogously to the bounding parameters $\epsilon$ in \ref{subsec:fgsm}, the outputs of the final layer are bounded by applying a $\tanh$ activation functions and the axes corresponding to $\pt$, $\eta$, and $\phi$ are scaled by parameters $\rho_{\pt}$, $\rho_\eta$, and $\rho_\phi$, respectively.

@@

- Pythia versus Herwig as a standard

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.45\textwidth]{figures/FGSM_Significance.pdf}
\caption{Top: The relative discovery significance (true positive rate / square root of false positive rate) as a function of the classifier threshold for HL observables and LL observables before and after the FGSM attack described in Sec.~\ref{subsec:fgsm}.  }
\label{fig:FGSM2}
\end{figure}
